---
title: "Non-linear Modeling"
author: "Marco Rizk"
date: "14 December 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Non-linear Modeling

```{r}
#loading data
library(ISLR)
attach(Wage)
```

## Polynomial Regression and step function

```{r}
#fitting 4th degree polynomial model

lm.fit <- lm(wage~poly(age,4),data=Wage)
coef(summary(lm.fit))
```
```{r}
#another way of fitting a polynomial model
fit2a=lm(wage???age+I(age^2)+I(age^3)+I(age^4),data=Wage)
coef(fit2a)
```

We now create a grid of values for age at which we want predictions, and
then call the generic predict() function, specifying that we want standard
errors as well.

```{r}
agelims =range(age)
age.grid=seq(from=agelims [1],to=agelims [2])
preds=predict (lm.fit ,newdata =list(age=age.grid),se=TRUE)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
```


```{r}
par(mfrow=c(1,2),mar=c(4.5,4.5,1,1) ,oma=c(0,0,4,0))
plot(age ,wage ,xlim=agelims ,cex =.5,col=" darkgrey ")
title(" Degree -4 Polynomial ",outer=T)
lines(age.grid ,preds$lm.fit ,lwd=2,col="blue")
matlines (age.grid ,se.bands ,lwd=1, col=" blue",lty=3)
```
Here the mar and oma arguments allow us to control the margins
of the plot, and the title() function creates a figure title that spans both title() subplots.

### using Anova to determine degree of polynomial

We use the anova() function, which performs an anova() analysis of variance (ANOVA, using an F-test) in order to test the null analysis of
hypothesis that a model M1 is sufficient to explain the data against the variance
alternative hypothesis that a more complex model M2 is required. In order
to use the anova() function, M1 and M2 must be nested models: the
predictors in M1 must be a subset of the predictors in M2. In this case,
we fit five different models and sequentially compare the simpler model to
the more complex model.

```{r}
#testing different degrees polynomial models
fit.1=lm(wage~age ,data=Wage)
fit.2=lm(wage~poly(age ,2),data=Wage)
fit.3=lm(wage~poly(age ,3),data=Wage)
fit.4=lm(wage~poly(age ,4),data=Wage)
fit.5=lm(wage~poly(age ,5),data=Wage)
anova(fit.1,fit.2,fit.3,fit.4,fit.5)
```

The p-value comparing the linear Model 1 to the quadratic Model 2 is
essentially zero (<10???15), indicating that a linear fit is not sufficient. Similarly the p-value comparing the quadratic Model 2 to the cubic Model 3
is very low (0.0017), so the quadratic fit is also insufficient. The p-value
comparing the cubic and degree-4 polynomials, Model 3 and Model 4, is approximately 5 % while the degree-5 polynomial Model 5 seems unnecessary
because its p-value is 0.37. Hence, either a cubic or a quartic polynomial
appear to provide a reasonable fit to the data, but lower- or higher-order
models are not justified.


### using cross validation to determine degree of polynomial

```{r}
fit=glm(I(wage >250)???poly(age ,4),data=Wage , family=binomial )
 pfit=exp(preds$fit )/(1+exp(preds$fit ))
se.bands.logit = cbind(preds$fit +2* preds$se.fit , preds$fit -2*
preds$se.fit)
se.bands = exp(se.bands.logit)/(1+exp(se.bands.logit))
```

```{r}
plot(age ,I(wage >250),xlim=agelims ,type="n",ylim=c(0,.2))
points(jitter(age), I((wage >250)/5),cex=.5,pch ="|",
col="darkgrey ")
lines(age.grid ,pfit ,lwd=2, col ="blue")
matlines (age.grid ,se.bands ,lwd=1, col=" blue",lty=3)
```

  We have drawn the age values corresponding to the observations with wage
values above 250 as gray marks on the top of the plot, and those with wage
values below 250 are shown as gray marks on the bottom of the plot. We
used the jitter() function to jitter the age values a bit so that observations
jitter() with the same age value do not cover each other up. This is often called a
rug plot. rug plot
In order to fit a step function, as discussed in Section 7.2, we use the
cut() function

```{r}
table(cut(age ,4))
fit=lm(wage???cut(age ,4),data=Wage)
coef(summary (fit))
```

Here cut() automatically picked the cutpoints at 33.5, 49, and 64.5 years
of age. We could also have specified our own cutpoints directly using the
breaks option. The function cut() returns an ordered categorical variable;
the lm() function then creates a set of dummy variables for use in the regression. The age<33.5 category is left out, so the intercept coefficient of
$94,160 can be interpreted as the average salary for those under 33.5 years
of age, and the other coefficients can be interpreted as the average additional salary for those in the other age groups. We can produce predictions
and plots just as we did in the case of the polynomial fit.


## splines

In order to fit regression splines in R, we use the splines library. In Section
7.4, we saw that regression splines can be fit by constructing an appropriate
matrix of basis functions. The bs() function generates the entire matrix of bs() basis functions for splines with the specified set of knots. By default, cubic
splines are produced. Fitting wage to age using a regression spline is simple:

```{r}
library(splines)
fit=lm(wage???bs(age ,knots=c(25,40,60) ),data=Wage)
pred=predict (fit ,newdata =list(age=age.grid),se=T)
plot(age ,wage ,col="gray")
lines(age.grid ,pred$fit ,lwd=2)
lines(age.grid ,pred$fit +2*pred$se ,lty="dashed")
lines(age.grid ,pred$fit -2*pred$se ,lty="dashed")
```
Here we have prespecified knots at ages 25, 40, and 60. This produces a
spline with six basis functions. (Recall that a cubic spline with three knots
has seven degrees of freedom; these degrees of freedom are used up by an
intercept, plus six basis functions.) We could also use the df option to
produce a spline with knots at uniform quantiles of the data.

```{r}
dim(bs(age ,knots=c(25,40,60) ))
dim(bs(age ,df=6))
attr(bs(age ,df=6) ,"knots")
```

In this case R chooses knots at ages 33.8, 42.0, and 51.0, which correspond
to the 25th, 50th, and 75th percentiles of age. The function bs() also has
a degree argument, so we can fit splines of any degree, rather than the
default degree of 3 (which yields a cubic spline).
In order to instead fit a natural spline, we use the ns() function. Here ns() we fit a natural spline with four degrees of freedom.

```{r}

fit2=lm(wage???ns(age ,df=4),data=Wage)
pred2=predict (fit2 ,newdata=list(age=age.grid),se=T)
{lines(age.grid , pred2$fit ,col="red",lwd=2)}
```
As with the bs() function, we could instead specify the knots directly using
the knots option.
```{r}
plot(age ,wage ,xlim=agelims ,cex =.5,col=" darkgrey ")
title("Smoothing Spline ")
fit=smooth.spline(age ,wage ,df=16)
fit2=smooth.spline (age ,wage ,cv=TRUE)
fit2$df
lines(fit ,col="red",lwd =2)
lines(fit2 ,col="blue",lwd=2)
legend ("topright ",legend=c("16 DF" ,"6.8 DF"),
col=c("red","blue"),lty=1,lwd=2, cex =.8)
```
Notice that in the first call to smooth.spline(), we specified df=16. The
function then determines which value of ?? leads to 16 degrees of freedom. In
the second call to smooth.spline(), we select the smoothness level by crossvalidation; this results in a value of ?? that yields 6.8 degrees of freedom.
In order to perform local regression, we use the loess() function.

```{r}
plot(age ,wage ,xlim=agelims ,cex =.5,col=" darkgrey ")
title("Local Regression ")
fit=loess(wage???age ,span=.2,data=Wage)
fit2=loess(wage???age ,span=.5,data=Wage)
lines(age.grid ,predict (fit ,data.frame(age=age.grid)),
col="red",lwd=2)
lines(age.grid ,predict (fit2 ,data.frame(age=age.grid)),
col="blue",lwd=2)
legend ("topright ",legend=c("Span=0.2"," Span=0.5"),
col=c("red","blue"),lty=1,lwd=2, cex =.8)
```
Here we have performed local linear regression using spans of 0.2 and 0.5:
that is, each neighborhood consists of 20 % or 50 % of the observations. The
larger the span, the smoother the fit. The locfit library can also be used
for fitting local regression models in R.

## GAMs

We now fit a GAM to predict wage using natural spline functions of year
and age, treating education as a qualitative predictor, as in (7.16). Since
this is just a big linear regression model using an appropriate choice of
basis functions, we can simply do this using the lm() function.

```{r}
 gam1=lm(wage???ns(year ,4)+ns(age ,5)+education ,data=Wage)
```

We now fit the model (7.16) using smoothing splines rather than natural
splines. In order to fit more general sorts of GAMs, using smoothing splines
or other components that cannot be expressed in terms of basis functions
and then fit using least squares regression, we will need to use the gam
library in R.
The s() function, which is part of the gam library, is used to indicate that s() we would like to use a smoothing spline. We specify that the function of
year should have 4 degrees of freedom, and that the function of age will
have 5 degrees of freedom. Since education is qualitative, we leave it as is,
and it is converted into four dummy variables. We use the gam() function in gam() order to fit a GAM using these components. All of the terms in (7.16) are
fit simultaneously, taking each other into account to explain the response.
```{r}
install.packages("gam")
```

```{r}
library (gam)
gam.m3=gam(wage???s(year ,4)+s(age ,5)+education ,data=Wage)
```

```{r}
par(mfrow=c(1,3))
plot(gam.m3, se=TRUE ,col =" blue")
```
The generic plot() function recognizes that gam2 is an object of class gam,
and invokes the appropriate plot.gam() method. Conveniently, even though plot.gam() gam1 is not of class gam but rather of class lm, we can still use plot.gam()
on it. Figure 7.11 was produced using the following expression:

```{r}
plot.gam(gam1 , se=TRUE , col="red")
```

Notice here we had to use plot.gam() rather than the generic plot()
function.
In these plots, the function of year looks rather linear. We can perform a
series of ANOVA tests in order to determine which of these three models is
best: a GAM that excludes year (M1), a GAM that uses a linear function
of year (M2), or a GAM that uses a spline function of year (M3).

```{r}
gam.m1=gam(wage???s(age ,5)+education ,data=Wage)
gam.m2=gam(wage???year+s(age ,5)+education ,data=Wage)
anova(gam.m1,gam.m2,gam.m3,test="F")
```

We find that there is compelling evidence that a GAM with a linear function of year is better than a GAM that does not include year at all
(p-value = 0.00014). However, there is no evidence that a non-linear function of year is needed (p-value = 0.349). In other words, based on the results
of this ANOVA, M2 is preferred.
The summary() function produces a summary of the gam fit

```{r}
summary (gam.m3)
```
The p-values for year and age correspond to a null hypothesis of a linear
relationship versus the alternative of a non-linear relationship. The large
p-value for year reinforces our conclusion from the ANOVA test that a linear function is adequate for this term. However, there is very clear evidence
that a non-linear term is required for age.
We can make predictions from gam objects, just like from lm objects,
using the predict() method for the class gam. Here we make predictions on
the training set.

```{r}
preds=predict (gam.m2,newdata =Wage)
```

  We can also use local regression fits as building blocks in a GAM, using
the lo() function.

```{r}
gam.lo=gam(wage???s(year ,df=4)+lo(age ,span =0.7)+education ,
data=Wage)
plot.gam(gam.lo, se=TRUE , col ="green")
```

Here we have used local regression for the age term, with a span of 0.7.
We can also use the lo() function to create interactions before calling the
gam() function. For example,

```{r}
gam.lo.i=gam(wage???lo(year ,age , span=0.5)+education ,
data=Wage)
```
fits a two-term model, in which the first term is an interaction between
year and age, fit by a local regression surface. We can plot the resulting
two-dimensional surface if we first install the akima package.
```{r}
install.packages("akima")
```

```{r}
library(akima)
plot(gam.lo.i)
```

In order to fit a logistic regression GAM, we once again use the I() function in constructing the binary response variable, and set family=binomial

```{r}
gam.lr=gam(I(wage >250)???year+s(age ,df=5)+education ,
family=binomial ,data=Wage)
par(mfrow=c(1,3))
plot(gam.lr,se=T,col="green ")
```
It is easy to see that there are no high earners in the <HS category:

```{r}
table(education ,I(wage >250))
```

Hence, we fit a logistic regression GAM using all but this category. This
provides more sensible results

```{r}
gam.lr.s=gam(I(wage >250)???year+s(age ,df=5)+education ,family=
binomial ,data=Wage , subset =(education !="1. < HS Grad"))
plot(gam.lr.s,se=T,col="green")
```

